
\chapter*{Declaration}
\addcontentsline{toc}{chapter}{Declaration and Impact statement}





This thesis is an account of research undertaken between February 2021 and  October 2021 at Research School of Biology, Joint College of Sciences, The Australian National University, Canberra, Australia.

Except where acknowledged, the material presented in this thesis is, to the best of my knowledge, original and has not been submitted in whole or part for a degree in any university.

\vspace{20mm}  % vertical space

\large
\begin{flushright}
Phuong Le \\
\today
\end{flushright}

\normalsize
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

First, I would like to thank my supervisor Gavin Huttley. Gavin started mentoring me two years ago when I took his course, Bioinformatics, with hardly any programming background. I then did a short project with him the following semester. Finally, I came back to undertake Honours with him this year. Over such a long period of time, my view remains the same, that I am privileged to be his student. He is one of the rare true bioinformaticians who have a profound understanding of all three domains: biology, computing and statistics. He genuinely cares about students' development, not just the outcome of our work. I always feel extremely assured when I talk with him, when he tells me to ``do good science'' and to ``do what we believe is right'' and that he has ``complete confidence in me''. His lab is a family where Katherine is apparently the bad twin and I am the evil twin. I cannot express how grateful I am for being instructed by Gavin. I hope my thesis and everything I do later on in life will make him proud.

Second, I would like to thank my co-supervisor, Cheng-Soon Ong. I have a tremendous respect for his expertise in machine learning and his superpower of communicating with biologists. While being a busy director of a machine learning platform at CSIRO, Cheng always makes time for me whenever I ping him on Slack. He is always concerned that I might be too stressed and that I do not take enough break. He connects me with his brilliant students and employees at Data61 so that I have even more people to hassle. Besides being a teacher, he gives plenty of great career and life advice. I wish I could quote everything he says, but let me just quote one: ``Be kind''.

I would also like to thank the three powerful individuals who control my fate, Aude Fahrer, Eric Stone and Sasha Mikheyev. I would like to acknowledge Eric, whose statistical mindset I find particularly eye-opening. With his experience, Eric's comments and advice have always been spot on. He has also referred me to his employees at the BDSI, whose expertise was in line with what I was doing. I cannot ask for a more helpful examiner. In addition, I would like to thank Aude, who I knew from her first year cancer lectures and her immunology course. Her input on the biological motivation for my project was invaluable. Communicating with her gives me the perfect chance to practice intellectual conversations with people outside my area.

Throughout the project, I had the opportunity to receive technical support from so many people. I wanted to thank anh\footnote{anh: Vietnamese for big brother} Mai Huy Hoang for patiently helping me with the ``computer stuff'' ever since we were high school students until now when he's finishing his degree at the prestigious Peking University. I wanted to thank my ge\footnote{ge: Chinese for big brother} Zixian Cai for his advice on everything, because conveniently for me, he is both a brilliant computer scientist and a senior international student of the same degree. I was also lucky to get help from Cheng's PhD student, Mengyan Zhang, who understands how to implement machine learning algorithms very well. She is kind enough to volunteer to give me advice that always points out both the strengths and weaknesses of my crazy ideas... I wanted to thank anh Hoang Danh Tai at the BDSI for sharing his machine learning experience and James Nichols for discussing Wasserstein distance with me. I wanted to thank Aaron Chuah at the JCSMR for sharing his wide knowledge of genomic databases. 

Thank you to the lovely people, besides my supervisors, who patiently tolerated my incredibly low speed of writing, read and edited different parts of my thesis, they are Soraya Zwahlen, Hanh Vo, Russell Tsuchida, Aaron Chuah, Andrea Do and Tai Hoang. Everyone was from a different background but all of them were so constructive with their feedback. I cannot imagine how much time and effort they have put in proofreading this, especially Soraya, whose comments were super detailed, super harsh but super encouraging at the same time. 

In terms of financial support, I would like to thank the ANU for awarding me the annual Terrell International Undergraduate Scholarships. On top of that, I would like to thank my past employers for letting me work for them to partly support myself when I moved to Australia, they are Susan Howitt, Mark Ellison, Quynh Nguyen, Johannes Zoller, Thuan Truong and Sao Berger. Susan Howitt, in particular, was another mentor of mine. She taught me during her genetics course, gave me honest advice in terms of career development and course selection. She was my referee when I transferred into the PhB. This year, I worked for her as a demonstrator and I can say that she was extremely caring as an employer. 

This year was ridiculously intense for me in many ways on top of the lockdown and I thank friends who were by my side to help me through it. Thank you to Katherine Caley, my lab mate. Throughout the year, Kath knows virtually everything I go through, she is there for me and she helps as much as she can. She enjoys our complaint sessions, she exchanges code with me and co-authors a few of my emails. Thanks to my roommate Chelsea Crew and landlords Brett and Barbara Still for their amazing kindness. Thanks to em\footnote{Vietnamese for little sister/brother} Han Do and Randolph Leong for being great friends and exchanging food with me. Thanks to my old roommate Hanh Vo for supporting me with useful materials even though she is now based in Tasmania (too bad you couldn't be my source of distraction as I was for you last year). Thanks to em Misa Le for being a great friend/roommate and for being the lucky charm who fixed the dodgy wifi just before my final seminar presentation, helped me decide the colour scheme for my thesis and graphics, forced me to enjoy my birthday just a week before thesis submission and pressed the submit button for me. Thanks to my aunt Le Kim Thuy, who has constant fear that I will starve myself to death and sends me heaps of delicious food. Thanks to my best friend Nguyen Thi Ha Anh, who always supports my superstition by being professionally supersitious herself. Thanks to Soraya Zwahlen, Hang Dang, Minh Bui, Thinh Ngo, Emily Tan, Sharwel Lei. And Maksim Lisau.

Another important person who got me through Honours was Spencer Whitney, who convened my Honours cohort. He is extremely experienced with students and he is always ready to help. He has a questionable sense of humour - he likes jokes that target my supervisor Gavin Huttley. I do too. 

Last but not least, I wanted to thank my family back in Vietnam, my mum Nguyen Thanh Thu, my dad Le Minh Quang and my sister Le Khanh Huong. My parents have been supporting me unconditionally. They are my main financial support, without which my study at ANU would have never been possible. But most importantly, they are my biggest emotional support that I can always rely on. When I am tired, I think of home. I guess there is no need to say anything further. 