\thispagestyle{empty}

\vspace*{3cm}

\begin{center}
All models are wrong, but some are useful \\ \medskip
--- George Box ---   
\end{center}

\vspace{2cm}

At the beginning of the year, I had a dream that I was alone back in my old house in Vietnam. The house was full of snakes and I was traumatised, but there was a tiny cat who talked with me and told me to run upstairs. So I ran upstairs, but on my way up, there were big black snakes sleeping after several steps. The cat taught me how to avoid the snakes while he dealt with them behind me. When I finally got upstairs, I went into a room, locked the door then jumped on the bed. Then I immediately discovered another snake on the floor. Unlike the other snakes, this one was green and translucent, and it was not sleeping. I realised that the cat was locked outside so he couldn't save me this time. Then the snake saw me. It jumped onto the bed to attack me. Luckily I shielded myself with the blanket, so I could push the snake down with my body, and the only thing that separated me from the snake was the blanket. The snake violently moved underneath and I had to use all my strength to fix it there. While I was freaked out and didn't know what to do, the cat somehow got inside. But even then, I was so close to the snake that the cat couldn't do anything - I was on my own. He threw at me a red book for me to figure it out. The book contained instructions on how to deal with the snake and `send it to the next life'. I remembered waking up at the time, carefully checking if there was any snake on my bed, thank God there wasn't any...

Honours was as intense as my dream. And because I am superstitious, I like to associate the snakes with Honours, and the cat with my supervisors, and everyone else who gave me advice along the way. Through the journey of Honours, I found that ideas are often generated and refined when I share them with other people. I also understand more what it means by ``If you want to go far, go together''. With that, I would like to share this \LaTeX\ format, which took me quite some time to make, with anyone who wanted it - contact me at  . It was inspired by André Miede's \href{https://bitbucket.org/amiede/classicthesis/wiki/Home}{\textit{A Classic Thesis Style}}, I think the code is a bit easier for beginners to tweak.